Iterative Monomorphisation

### 2025-06-26 iterative_monomorphisation.tex
cleanup
spellings

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}

@@

%These definitions depend on the precise form of the input formulae, which has been left undefined.
\(N\) is parameterised with \(\varphi\) because type variables are implicitly quantified at the formula level. The formula indicates the scope of type variables. This is not necessary for \(M\) since all the types it contains are monomorphic.

@@

To illustrate the low level version of the iterative algorithm, we consider the following initial problem:
\begin{itemize}
   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{p}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{p}\langle \alpha, \alpha \rangle(x, x) \land \sym{p}\langle \alpha, \beta\rangle(x, z)\)
\end{itemize}
%
\(M_{\text{new}}\) is initialised with \(\{\sym p\mapsto (\ty{int}, \ty{nat}); \sym p\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{\sym p\mapsto (\ty{list}(\alpha), \alpha); \sym p\mapsto (\alpha, \alpha); \sym p\mapsto (\alpha, \beta)\}\}\). Then we enter the main loop (assuming \Loop is at least 1). We iterate over each formula and call \FStep for each of them. Nothing happens for formula \(\langle1\rangle\) because it contains no type variables. For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.

@@

Finally, if the target language supports only nullary type construtors, the types of all monomorphic formulae, both new and old, are mangled.
%This step is unecessary if the target language allows type constructors.
In our example, two new monomorphic formulae are generated. The algorithm outputs
\begin{enumerate}
   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle3\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \begin{aligned}[t]
     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
     & \sym{p}\langle \ty{int}, \ty{int}\rangle(x, z)\end{aligned}\)
   \item[\(\langle4\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \begin{aligned}[t]
     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
     & \sym{p}\langle \ty{int}, \ty{nat}\rangle(x, z)\end{aligned}\)
\end{enumerate}

@@

Table \ref{e_settings} shows the impact of the options with which we call the E prover.
The point at which Zipperposition interupts its normal operation and begins the monomorphisation process is determined by the `E call point' parameter. It is expressed as a fraction of the total time alloted to Zipperposition.
The E timeout (in seconds) limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation. The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is. This likely explains the poor performance for greater values of the E call point.

>>>>

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encodings may result in incorrect characters.
%
\usepackage{graphicx}

@@

%These definitions depend on the precise form of the input formulae, which has been left undefined.
\(N\) is parametrised with \(\varphi\) because type variables are implicitly quantified at the formula level. The formula indicates the scope of type variables. This is not necessary for \(M\) since all the types it contains are monomorphic.

@@

To illustrate the low level version of the iterative algorithm, we consider the following (admittedly contrived) initial problem:
\begin{itemize}
   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{p}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{p}\langle \alpha, \alpha \rangle(x, x) \land \sym{p}\langle \alpha, \beta\rangle(x, z)\)
\end{itemize}
%
\(M_{\text{new}}\) is initialised with \(\{\sym p\mapsto (\ty{int}, \ty{nat}); \sym p\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{\sym p\mapsto (\ty{list}(\alpha), \alpha); \sym p\mapsto (\alpha, \alpha); \sym p\mapsto (\alpha, \beta)\}\}\). Then we enter the main loop (assuming \Loop is at least 1). We iterate over each formula and call \FStep for each of them. Nothing happens for formula \(\langle1\rangle\) because it contains no type variables. For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.

@@

Finally, if the target language supports only nullary type construtors, the types of all monomorphic formulae, both new and old, are mangled.
%This step is unnecessary if the target language allows type constructors.
In our example, two new monomorphic formulae are generated. The algorithm outputs
\begin{enumerate}
   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle3\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \begin{aligned}[t]
     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
     & \sym{p}\langle \ty{int}, \ty{int}\rangle(x, z)\end{aligned}\)
   \item[\(\langle4\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \begin{aligned}[t]
     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
     & \sym{p}\langle \ty{int}, \ty{nat}\rangle(x, z)\end{aligned}\)
\end{enumerate}

@@

Table \ref{e_settings} shows the impact of the options with which we call the E prover.
The point at which Zipperposition interrupts its normal operation and begins the monomorphisation process is determined by the `E call point' parameter. It is expressed as a fraction of the total time allotted to Zipperposition.
The E timeout (in seconds) limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation. The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is. This likely explains the poor performance for greater values of the E call point.
